SampleSizePlanner: A Tool to Estimate and Justify Sample Size for Two-Group Studies

### 2021-09-02 inst/app/www/about.Rmd
renaming, tidying, recalculating ssp results, debugging

The __SampleSizePlanner__ shiny application enables researchers to calculate the sample size with 9 different methods in app and provides code to use an additional method from R. The aim of the app is to make the various sample size calculating methods available for a wide range of researchers and propagate the transparent reporting of the chosen method. Therefore, a formatted template text containing the used parameters and the output of the calculations is available for download. This text can be complemented with the reasoning behind each decision.

### Contributors
- Marton Kovacs
- Don van Ravenzwaaij
- Rink Hoekstra
- Balazs Aczel

@@

If you have any questions or comments please write to Marton Kovacs to the following email address: marton.balazs.kovacs@gmail.com

Or submit an issue on the apps Github page: [https://github.com/marton-balazs-kovacs/SampleSizePlanner](https://github.com/marton-balazs-kovacs/SampleSizePlanner)

>>>>

The **SampleSizePlanner** shiny application enables researchers to calculate the sample size with 9 different methods in app and provides code to use an additional method from R. The aim of the app is to make the various sample size calculating methods available for a wide range of researchers and propagate the transparent reporting of the chosen method. Therefore, a formatted template text containing the used parameters and the output of the calculations is available for download. This text can be complemented with the reasoning behind each decision.

### Contributors

-   Marton Kovacs
-   Don van Ravenzwaaij
-   Rink Hoekstra
-   Balazs Aczel

The design of the app was done by Zsuzsa Szekely [(@szekely_Zsuzsa)](<https://twitter.com/szekely_zsuzsa>).

@@

If you have any questions or comments please write to Marton Kovacs to the following email address: [marton.balazs.kovacs@gmail.com](mailto:marton.balazs.kovacs@gmail.com)

Or submit an issue on the apps Github page: <https://github.com/marton-balazs-kovacs/SampleSizePlanner>

### 2021-09-02 inst/manuscript/manuscript.Rmd
renaming, tidying, recalculating ssp results, debugging

library("papaja")
library("formatR")
library("kableExtra")
devtools::install_github("marton-balazs-kovacs/SampleSizePlanner")
```

@@

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She believes that distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. For illustrative purposes, we will have Mary's hypothetical study goal depend on the earlier presented decisions to illustrate different scenarios for each sample size planning method. That means that in some of the upcoming examples, Mary expects a null effect and in others she expects a positive effect. We will also present a justification text based on Mary’s choices for each method.

@@

```{r powercurve, echo = FALSE, fig.cap = "(ref:powercurve-caption)", fig.align = "center",}
# Determine the sample sizes for each delta
curve_data <- SampleSizePlanner::ssp_power_curve(
  tpr = 0.8, delta = seq(0.1, 0.9, 0.01), max_n = 5000
  )
# Plot the power curve
SampleSizePlanner::plot_power_curve(
  delta = curve_data$delta, n1 = curve_data$n1, animated = FALSE
  )
```

>>>>

library("papaja")
library("ggplot2")
library("formatR")
library("kableExtra")
devtools::install_github("marton-balazs-kovacs/SampleSizePlanner")
```

@@

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She would like to test whether distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. Mary has good reason to be skeptical about the effectiveness of training on increasing performance as there are several studies questioning the existence of such effects [@owen_putting_2010; @simons_brain-training_2016].     For illustrative purposes, in some of the upcoming examples Mary expects a null effect and in others Mary expects a positive effect in order to highlight the different research scenarios for each sample size planning method. We will also present a justification text for each sample size planning method based on Mary’s choices described in the example research scenario for the given method.

@@

```{r powercurve, echo = FALSE, fig.cap = "(ref:powercurve-caption)", fig.align = "center",}
# Determine the sample sizes for each delta
curve_data <- SampleSizePlanner::ssp_power_curve(
  tpr = 0.8, delta = seq(0.1, 0.9, 0.01), max_n = 5000
  )
# Plot the power curve
SampleSizePlanner::plot_power_curve(
  delta = curve_data$delta, n1 = curve_data$n1, animated = FALSE
  )

# Save the plot in pdf
ggplot2::ggsave("fig2.pdf", plot = last_plot(), dpi = 300)
```

### 2021-09-02 inst/manuscript/manuscript.tex
renaming, tidying, recalculating ssp results, debugging

It is crucial to not just state how we determined our planned sample size but to also give the reader insight into the reasons behind our choices. In a recent overview, Lakens (2021) lists six types of general approaches to justify sample size in quantitative empirical studies: (1) Measure entire population; (2) Resource constraints; (3) A-priori power analysis; (4) Accuracy; (5) Heuristics; and (6) No justification. For the first approach, no quantitative justification is necessary; and for the second approach, the researcher has no freedom to increase the sample size. Power analysis, or more generally the estimation of true positive rate, is used when one plans to conduct hypothesis testing; accuracy justifications are used when one plans to conduct parameter estimation. Our tutorial mainly focuses on approaches two, three, and four, and is aimed at providing a hands-on approach for the mechanical part of the sample size determination (i.e., the calculation). For a deeper discussion of justification of these approaches, or for other approaches (i.e., using heuristics or not providing justification), we refer the reader to Lakens (2021).

@@

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She believes that distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. For illustrative purposes, we will have Mary's hypothetical study goal depend on the earlier presented decisions to illustrate different scenarios for each sample size planning method. That means that in some of the upcoming examples, Mary expects a null effect and in others she expects a positive effect. We will also present a justification text based on Mary's choices for each method.

@@

TOST is a frequentist equivalence testing approach that adopts two one-sided hypotheses to designate an interval hypothesis (Schuirmann, 1987). The lower and upper boundaries of the interval are determined by the equivalence band (i.e.~SESOI) around the expected population effect size (e.g., 0). Lakens, Scheel, and Isager (2018) lists several methods that can be used to determine the SESOI . In case of TOST, the two null hypotheses state that the effect size is equal to the lower and upper equivalence band values, whereas the alternative hypotheses state that the effect size is significantly smaller than the upper equivalence band value and significantly larger than the lower equivalence band value. In case both one-sided tests reject the null-hypothesis at a given significance level, the group means are considered to be practically equivalent. See Lakens, Scheel, and Isager (2018), for further reading.

@@

\leavevmode\hypertarget{ref-ioannidis_why_2005}{}%
Ioannidis, J. P. (2005). Why most published research findings are false. \emph{{PLoS} Medicine}, \emph{2}(8), e124.

\leavevmode\hypertarget{ref-kelley_sample_2006}{}%
Kelley, K., \& Rausch, J. R. (2006). Sample size planning for the standardized mean difference: Accuracy in parameter estimation via narrow confidence intervals. \emph{Psychological Methods}, \emph{11}(4), 363--385.

@@

\leavevmode\hypertarget{ref-kruschke_rejecting_2018}{}%
Kruschke, J. K. (2018). Rejecting or accepting parameter values in bayesian estimation. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 270--280.

\leavevmode\hypertarget{ref-lakens_sample_2021}{}%
Lakens, D. (2021). Sample size justification. Retrieved from \url{https://doi.org/10.31234/osf.io/9d3yf}

\leavevmode\hypertarget{ref-lakens_equivalence_2018}{}%
Lakens, D., Scheel, A. M., \& Isager, P. M. (2018). Equivalence testing for psychological research: A tutorial. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 259--269.

@@

\leavevmode\hypertarget{ref-maxwell_persistence_2004}{}%
Maxwell, S. E. (2004). The persistence of underpowered studies in psychological research: Causes, consequences, and remedies. \emph{Psychological Methods}, \emph{9}(2), 147--163.

\leavevmode\hypertarget{ref-morey_bayes_2011}{}%
Morey, R. D., \& Rouder, J. N. (2011). Bayes factor approaches for testing interval null hypotheses. \emph{Psychological Methods}, \emph{16}(4), 406--419.

\leavevmode\hypertarget{ref-van_ravenzwaaij_bayes_2019}{}%
Ravenzwaaij, D. van, Monden, R., Tendeiro, J. N., \& Ioannidis, J. P. (2019). Bayes factors for superiority, non-inferiority, and equivalence designs. \emph{{BMC} Medical Research Methodology}, \emph{19}(1), 1--12.

@@

\leavevmode\hypertarget{ref-schonbrodt_bayes_2018}{}%
Schönbrodt, F. D., \& Wagenmakers, E.-J. (2018). Bayes factor design analysis: Planning for compelling evidence. \emph{Psychonomic Bulletin \& Review}, \emph{25}(1), 128--142.

\leavevmode\hypertarget{ref-schonbrodt_sequential_2017}{}%
Schönbrodt, F. D., Wagenmakers, E.-J., Zehetleitner, M., \& Perugini, M. (2017). Sequential hypothesis testing with bayes factors: Efficiently testing mean differences. \emph{Psychological Methods}, \emph{22}(2), 322--339.

@@

>>>>

It is crucial to not just state how we determined our planned sample size but to also give the reader insight into the reasons behind our choices. In a recent overview, Daniel Lakens (2021) lists six types of general approaches to justify sample size in quantitative empirical studies: (1) Measure entire population; (2) Resource constraints; (3) A-priori power analysis; (4) Accuracy; (5) Heuristics; and (6) No justification. For the first approach, no quantitative justification is necessary; and for the second approach, the researcher has no freedom to increase the sample size. Power analysis, or more generally the estimation of true positive rate, is used when one plans to conduct hypothesis testing; accuracy justifications are used when one plans to conduct parameter estimation. Our tutorial mainly focuses on approaches two, three, and four, and is aimed at providing a hands-on approach for the mechanical part of the sample size determination (i.e., the calculation). For a deeper discussion of justification of these approaches, or for other approaches (i.e., using heuristics or not providing justification), we refer the reader to Daniel Lakens (2021).

@@

Throughout, we will use the example story of Mary the educational psychologist. Mary has come up with a new set of games that challenge spatial insight. She would like to test whether distributed and targeted engagement with these games for a period of six months for children in the age range of 8 to 12 will lead to lasting improvements on their IQ score as measured through Raven's progressive matrices test (population mean 100, population SD 15). Mary collects data for a control sample that gets regular education and for an experimental sample and plans to compare those samples. Mary has good reason to be skeptical about the effectiveness of training on increasing performance as there are several studies questioning the existence of such effects (Owen et al., 2010; Simons et al., 2016). For illustrative purposes, in some of the upcoming examples Mary expects a null effect and in others Mary expects a positive effect in order to highlight the different research scenarios for each sample size planning method. We will also present a justification text for each sample size planning method based on Mary's choices described in the example research scenario for the given method.

@@

TOST is a frequentist equivalence testing approach that adopts two one-sided hypotheses to designate an interval hypothesis (Schuirmann, 1987). The lower and upper boundaries of the interval are determined by the equivalence band (i.e.~SESOI) around the expected population effect size (e.g., 0). Daniël Lakens, Scheel, and Isager (2018) lists several methods that can be used to determine the SESOI . In case of TOST, the two null hypotheses state that the effect size is equal to the lower and upper equivalence band values, whereas the alternative hypotheses state that the effect size is significantly smaller than the upper equivalence band value and significantly larger than the lower equivalence band value. In case both one-sided tests reject the null-hypothesis at a given significance level, the group means are considered to be practically equivalent. See Daniël Lakens, Scheel, and Isager (2018), for further reading.

@@

\leavevmode\hypertarget{ref-ioannidis_why_2005}{}%
Ioannidis, J. P. (2005). Why most published research findings are false. \emph{PLoS Medicine}, \emph{2}(8), e124.

\leavevmode\hypertarget{ref-kelley_sample_2006}{}%
Kelley, K., \& Rausch, J. R. (2006). Sample size planning for the standardized mean difference: {Accuracy} in parameter estimation via narrow confidence intervals. \emph{Psychological Methods}, \emph{11}(4), 363.

@@

\leavevmode\hypertarget{ref-kruschke_rejecting_2018}{}%
Kruschke, J. K. (2018). Rejecting or accepting parameter values in {Bayesian} estimation. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 270--280.

\leavevmode\hypertarget{ref-lakens_sample_2021}{}%
Lakens, Daniel. (2021). Sample {Size} {Justification}.

\leavevmode\hypertarget{ref-lakens_equivalence_2018}{}%
Lakens, Daniël, Scheel, A. M., \& Isager, P. M. (2018). Equivalence testing for psychological research: {A} tutorial. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 259--269.

@@

\leavevmode\hypertarget{ref-maxwell_persistence_2004}{}%
Maxwell, S. E. (2004). The persistence of underpowered studies in psychological research: Causes, consequences, and remedies. \emph{Psychological Methods}, \emph{9}(2), 147.

\leavevmode\hypertarget{ref-morey_bayes_2011}{}%
Morey, R. D., \& Rouder, J. N. (2011). Bayes factor approaches for testing interval null hypotheses. \emph{Psychological Methods}, \emph{16}(4), 406.

\leavevmode\hypertarget{ref-owen_putting_2010}{}%
Owen, A. M., Hampshire, A., Grahn, J. A., Stenton, R., Dajani, S., Burns, A. S., \ldots{} Ballard, C. G. (2010). Putting brain training to the test. \emph{Nature}, \emph{465}(7299), 775--778.

\leavevmode\hypertarget{ref-van_ravenzwaaij_bayes_2019}{}%
Ravenzwaaij, D. van, Monden, R., Tendeiro, J. N., \& Ioannidis, J. P. (2019). Bayes factors for superiority, non-inferiority, and equivalence designs. \emph{BMC Medical Research Methodology}, \emph{19}(1), 1--12.

@@

\leavevmode\hypertarget{ref-schonbrodt_bayes_2018}{}%
Schönbrodt, F. D., \& Wagenmakers, E.-J. (2018). Bayes factor design analysis: {Planning} for compelling evidence. \emph{Psychonomic Bulletin \& Review}, \emph{25}(1), 128--142.

\leavevmode\hypertarget{ref-schonbrodt_sequential_2017}{}%
Schönbrodt, F. D., Wagenmakers, E.-J., Zehetleitner, M., \& Perugini, M. (2017). Sequential hypothesis testing with {Bayes} factors: {Efficiently} testing mean differences. \emph{Psychological Methods}, \emph{22}(2), 322.

@@

\leavevmode\hypertarget{ref-simons_brain-training_2016}{}%
Simons, D. J., Boot, W. R., Charness, N., Gathercole, S. E., Chabris, C. F., Hambrick, D. Z., \& Stine-Morrow, E. A. (2016). Do {``brain-training''} programs work? \emph{Psychological Science in the Public Interest}, \emph{17}(3), 103--186.
